Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-14 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
FH Table  CH 8
Incorp CF comments in Ch1

\acknowledgments{
I would like to thank Prof.~Fleischmann, my senior advisor with the University of Canterbury, as well as, Dr.~Kevin McGrattan, and Dr.~Craig Weinschenk from the National Institute of Standards and Technology for their guidance, assistance, and support throughout this study.

@@

Last but not least, I thank my wife, Liz, for her patience and support of my pursuit of higher education throughout our 37 years of marriage.  I would also like to thank my children; Catherine, Stephanie, Michael, Christina, and Veronica for ``buying me time'' to work on my studies.}

@@

List of Symbols
\begin{tabbing}
\hspace{1.5in} \= \\
$c_{p}$ \> specific heat \\
$k$ \> thermal conductivity \\
$\chi_r$ \> radiative fraction \\
discusion$\dot{q}''$ \> heat flux per unit area kW/m$^2$ \\
$\dot{Q}$ \> heat release rate (kW) \\ 
$R$ \> distance from the point source to the target (m) \\
$\rho$ \> density \\
\end{tabbing}

@@

\subsection{The Use of Fire Models in Fire Investigation}
Fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigator’s hypothesis on origin and cause~\cite{Sutula:2001}.  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs)~\cite{NUREG_1805}. These spreadsheets incorporate references to the source documents and experimental work upon which the correlations are based. NRC tasked NIST with conducting a verification and validation study of these correlations. Overholt developed the NIST verification and validation study~\cite{Overholt:2014}. Not all of the correlations in the FDTs were included in the study, because an insufficient amount of reliable data is available to conduct validation.  The correlations for flame height and time to ignition were not included in the validation study.

@@

The majority of intentional fires in structures have occurred in residential occupancies, therefore this study focused on fire patterns on painted gypsum wallboard which is a common interior finish for residential occupancies.  A common height for a residential occupancies is 2.4~m in height.  With a wall height of 2.4~m, these experiments used small fires with flame heights of approximately 1.2~m.  The resulting patterns on the gypsum board wall would be representative of a pre-flashover fire pattern.

@@

Replicate experiments exposing gypsum wallboard to the fires were conducted in order to examine the repeatability of the fire patterns.  Correlations between the fire characteristics and the fire patterns will be examined.

@@

FDS~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide} were used to examine the ability to numerically reproduce the source fires and the instrumented wall experiments. These simulation tools provide results in the form of visual field data.  In other words they provide results in the same format that the investigator would see or document the fire pattern.  Comparison between the experiments and the model may provide valuable validation data for the model.

@@

For the moment let's step into the shoes of a fire investigator and examine a potential investigation scenario.  It is 8:30 in the morning on a cool fall morning and you are being called to a fire scene.  As you prepare to leave for the fire scene, you begin to collect information on the fire as you listen to your department radio.  You have the address.  You start to develop an image in your mind about the location of the home involved in the fire. The fire occurred in area of older townhomes.  The socio-economics of the neighborhood are such the the people that live there have many life struggles. Enroute you learn that the fire department rescued a mother and child from the home and they are being transported to the hospital.  The husband and father also lives at that address but is unaccounted for at this time.  Upon arrival you have a conversation with the police and fire officers on the scene.  You are provided with information regarding what the first firefighters saw on the scene; ``all of the doors and windows were closed, and smoke was leaking from around the front door and living room window.  The front door was forced and a hose stream was applied to the fire in the living room.  Fire knockdown was rapid.  The mother and child were rescued from a bedroom upstairs.'' You thank the fire officer and get contact information for a follow-up if necessary.  As you wait for the fire department to complete their operations inside, the police on scene let you know that the address is familiar to them for domestic disturbance calls for the husband and wife fighting but no charges have been filed.  They are looking for the husband and have sent a patrol car to his place of work.

As you begin to examine the scene, you walk around the exterior and note that on the front of the end unit townhouse, there is fire damage on the exterior above the front door opening that extends across the ceiling of the porch.  There is some thermal damage and smoke deposits on the front siding above the porch.  The outer pane of the living room window is, for the most part, still intact. You notice that three of the four windows on the second floor have been removed.  On ``side D'' of the townhome there are no openings and no smoke or fire damage is present.  On the rear or ``side C'' of the house the door and windows are intact but open.  There are no smoke stains or thermal damage around any of the openings on the rear. No sign of smoke alarms in the home.

You take some notes and photograph the exterior of the structure, the fire department is repacking their equipment and are preparing to leave when you enter through the front doorway.  The living room has extensive fire damage, as shown in Figure~\ref{EXT_INTLR}.  Looking to your right, two chairs in front of the living room window have some thermal damage.  There are is a clear horizontal line of demarcation on the large wood TV stand in the corner.  As your eyes move from the corner back toward the rear of the townhome and the kitchen, you notice that the elevation of the line of demarcation is increasing.  This may be an indicator that the origin of the fire is closer to the front of the living room.  Between the two wood cabinets on the ``D wall'' of the living room there is a wood framed sofa.  The cushions have thermal damage, from pyrolysis.  It is not clear that they were on fire as there is no thermal damage on the wall directly above the top of the sofa cushions that would indicate the presence of a thermal plume.  Still standing in the entry doorway, you turn your attention to the remains of an upholstered sofa and chair.  The wall behind the sofa and chair has extensive thermal damage.  The wall extends from the opening to the dining/kitchen area of the house to the open stairway that goes to the upper level.  You move to the dining area and kitchen area.  There is plenty of smoke damage high in the space, but there are no indications of fire damage due to combustion.  You return to the living room and walk up the stairs.  There is evidence of paint and paper burned off of the gypsum wallboard in the stairway.  The thermal damage does not extend beyond the top of the stairs.  There is smoke damage in the upstairs hallway and in the front bedrooms. The further you move from the stairs the less thermal damage you find.

You head back down the stairs to further examine the living room which appears to be the room of fire origin. Your are focusing on the area with the most damage, the area with the sofa and chair.  As you are looking toward the ``B wall'' of the living room you note that the right side of the chair is badly burned and it is burned from the floor level up.  The same is true from the left side of the sofa.  In fact, the left side of the sofa has more fire damage to the wood frame of the sofa than the right side of the sofa frame.  The corner formed by the wall behind the chair and the wall of the stairway has a ``V pattern'' burned into it, from the floor and extending up toward the ceiling.  As you look at the debris surrounding the chair, you notice the remains of an electric heater.  The burned wire from the heater extends toward the wall outlet behind the sofa. Figure~\ref{EXT_LRDetails} shows some of these details.

You have now developed your first hypothesis about the origin of the fire.  A radiant style, electric heater was turned on.  It was too close to the furniture and ignited the fabric on the sofa or chair.  You collect the remains of the heater for further examination to have a lab determine if the heater switch was in the ``on'' position.  Now you need to start testing your hypothesis.  Is everything consistent with the other fire damage?

There are different damage patterns behind the sofa.  On the right side there is fire damage on the wall behind the right side back cushion of the sofa. The damaged area is distinct from the other fire damage that exists on the wall behind the sofa. There are pieces of charred newspaper near the right arm of the sofa.  The burn mark on the top of the wood coffee table looks like an arrow that is pointing toward the corner.  The char on the legs of the coffee table is very directional and consistent with the heat coming from the corner.  Looking at the inside area of the sofa, you see that there is charred wood on the inside surface of the sofa frame while the outer portions of the wood frame are not charred.  This would indicate that the fire was inside the sofa at some point.  Further is appears that all of the foam cushioning is burnt off of the right side of the sofa, while there is a piece of polyurethane foam cushion on the left side of the sofa.  However that foam could have gotten moved during the fire suppression.

The chair has more unburned foam cushioning then the sofa does and the seat cushion is still in place with it's distinctive shape, so you know that cushion has not been moved.  The cushion was heated from the top down.  Perhaps the construction of the chair slowed the flame spread through the chair from the bottom up, if the heater was the ignition source.  Or the fire started in the sofa and the radiation from the flames and hot gas layer spread the fire to the chair  The heater still looks good as a source however now you have some fire damage that may not be consistent with the heater as the origin.  Besides interviews what else can you do to eliminate the inconsistencies?   The fire investigators' challenge we be reexamined in Chapter IX.

@@

\begin{table}[h]
  \small
  \centering
  \begin{tabular}{|c|c|c|c|} \hline 
Distance from Wall 		&  Flame Height 			  	& RI=5       		& RI=10      		  \\
    					&  Range          				&  (80~mm grid)  	&  (40~mm grid)  	  \\
    (m)					&  (m) 							& (m) 				& (m) 				  \\ 
\hline    0.61			&  0.73 to 1.14         		& 1.02 to 1.27      & 1.xx to 1.28         \\
\hline    0.31			&								&					&					    \\
\hline    0.15			&								&					&						\\
\hline	  0.0			&								&					&  1.xx to 1.46 	  \\
\hline
  \end{tabular}
  \caption[Comparison of estimations of flame height from the natural gas instrumented wall experiments and FDS simulations with RIs of 5 and 10.]{Comparison of estimations of flame height from the natural gas instrumented wall experiments and FDS simulations with RIs of 5, and 10.}
  \label{tab:FDS_FH_IWNG}
\end{table}

>>>>

\acknowledgments{
I would like to thank Prof.~Fleischmann, my senior advisor with the University of Canterbury, Dr.~Kevin McGrattan from the National Institute of Standards and Technology, and Dr.~Craig Weinschenk from the UL Firefigther Safety Research Institute for their guidance, assistance, and support throughout this study.

@@

Last but not least, I thank my wife, Liz, for her patience and support of my pursuit of higher education throughout our 38 years of marriage.  I would also like to thank my children; Catherine, Stephanie, Michael, Christina, and Veronica for ``buying me time'' to work on my studies.}

@@

List of Symbols
\begin{tabbing}
\hspace{1.5in} \= \\
$c_{p}$ \> specific heat \\
$k$ \> thermal conductivity \\
$\chi_r$ \> radiative fraction \\
$\dot{q}''$ \> heat flux per unit area kW/m$^2$ \\
$\dot{Q}$ \> heat release rate (kW) \\ 
$R$ \> distance from the point source to the target (m) \\
$\rho$ \> density \\
\end{tabbing}

@@

\subsection{The Use of Fire Models in Fire Investigation}
Fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigator’s hypothesis on origin and cause~\cite{Sutula:2001}.  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs)~\cite{NUREG_1805}. These spreadsheets incorporate references to the source documents and experimental work upon which the correlations are based. NRC tasked NIST with conducting a verification and validation study of these correlations. Overholt developed the NIST verification and validation study~\cite{Overholt:2014}. Not all of the correlations in the FDTs were included in the study, because a sufficient amount of reliable data was not available to conduct a proper validation.  The correlations for flame height and time to ignition were not included in the validation study.

@@

The majority of intentional fires in structures have occurred in residential occupancies, therefore this study focused on fire patterns on painted gypsum wallboard which is a common interior finish for residential occupancies.  A common height for a residential occupancies is 2.4~m.  With a wall height of 2.4~m, these experiments used small fires with flame heights of approximately 1.2~m.  The resulting patterns on the gypsum board wall would be representative of a pre-flashover fire pattern.

@@

Replicate experiments exposing gypsum wallboard to the fires were conducted in order to examine the repeatability of the fire patterns.  Correlations between the fire characteristics and the fire patterns were be examined.

@@

FDS~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide} were used to examine the ability to numerically reproduce the source fires and the instrumented wall experiments. These simulation tools provide results in the form of visual field data.  In other words they provide results in the same format that the investigator would see or document the fire pattern.  Comparison between the experiments and the model provide valuable validation data for the model.

@@

N.B. This section may be considered atypical for an engineering thesis but it is included here for two reasons.  Firstly, to provide a narrative description for readers that may be unfamiliar with how fire patterns are routinely relied on in fire investigations.  The second is to put this work into an applied context that is familiar to fire investigators that may be un accustomed to reading detailed research as found in this thesis.

For the moment let's step into the shoes of a fire investigator and examine a potential investigation scenario.  It is 8:30 on a cool fall morning and you are being called to a fire scene.  As you prepare to leave for the fire scene, you begin to collect information on the fire as you listen to your department radio.  You have the address.  You start to develop an image in your mind about the location of the home involved in the fire. The fire occurred in area of older townhomes.  The socio-economics of the neighborhood are such that the people that live there have many life struggles. Enroute you learn that the fire department rescued a mother and child from the home and they are being transported to the hospital.  The husband and father also lives at that address but is unaccounted for at this time.  Upon arrival you have a conversation with the police and fire officers on the scene.  You are provided with information regarding what the first firefighters saw on the scene; ``all of the doors and windows were closed, and smoke was leaking from around the front door and living room window.  The front door was forced and a hose stream was applied to the fire in the living room.  Fire knockdown was rapid.  The mother and child were rescued from a bedroom upstairs.'' You thank the fire officer and get contact information for a follow-up if necessary.  As you wait for the fire department to complete their operations inside, the police on scene let you know that the address is familiar to them for domestic disturbance calls for the husband and wife fighting but no charges have been filed.  They are looking for the husband and have sent a patrol car to his place of work.

As you begin to examine the scene, you walk around the exterior and note that on the front of the end unit townhouse, there is fire damage on the exterior above the front door opening that extends across the ceiling of the porch.  There is some thermal damage and smoke deposits on the front siding above the porch.  The outer pane of the living room window is, for the most part, still intact. You notice that three of the four windows on the second floor have been removed.  On ``side D'' of the townhome there are no openings and no smoke or fire damage is present.  On the rear or ``side C'' of the house the door and windows are intact but open.  There are no smoke stains and no thermal damage around any of the openings on the rear. No sign of smoke alarms in the home.

You take some notes and photograph the exterior of the structure, the fire department is repacking their equipment and are preparing to leave when you enter through the front doorway.  The living room has extensive fire damage, as shown in Figure~\ref{EXT_INTLR}.  Looking to your right, two chairs in front of the living room window have some thermal damage.  There is a clear horizontal line of demarcation on the large wood TV stand in the corner.  As your eyes move from the corner back toward the rear of the townhome and the kitchen, you notice that the elevation of the line of demarcation is increasing.  This may be an indicator that the origin of the fire is closer to the front of the living room.  Between the two wood cabinets on the ``D wall'' of the living room there is a wood framed sofa.  The cushions have thermal damage, from pyrolysis.  It is not clear that they were on fire as there is no thermal damage on the wall directly above the top of the sofa cushions that would indicate the presence of a thermal plume.  Still standing in the entry doorway, you turn your attention to the remains of an upholstered sofa and chair.  The wall behind the sofa and chair has extensive thermal damage.  The wall extends from the opening to the dining/kitchen area of the house to the open stairway that goes to the upper level.  You move to the dining area and kitchen area.  There is plenty of smoke damage high in the space, but there are no indications of fire damage due to combustion.  You return to the living room and walk up the stairs.  There is evidence of paint and paper burned off of the gypsum wallboard in the stairway.  The thermal damage does not extend beyond the top of the stairs.  There is smoke damage in the upstairs hallway and in the front bedrooms. The further you move from the stairs the less thermal damage you find.

You head back down the stairs to further examine the living room which appears to be the room of fire origin. Your are focusing on the area with the most damage, the area with the sofa and chair.  You think that this may be the area of origin.

As you are looking toward the ``B wall'' of the living room you note that the right side of the chair is badly burned and it is burned from the floor level up.  The same is true from the left side of the sofa.  In fact, the left side of the sofa has more fire damage to the wood frame of the sofa than the right side of the sofa frame.  The corner formed by the wall behind the chair and the wall of the stairway has a ``V pattern'' burned into it, from the floor and extending up toward the ceiling.  As you look at the debris surrounding the chair, you notice the remains of an electric heater.  The burned wire from the heater extends toward the wall outlet behind the sofa. Figure~\ref{EXT_LRDetails} shows some of these details.

You have now developed your first hypothesis about the cause of the fire.  A radiant style, electric heater was turned on.  It was too close to the furniture and ignited the fabric on the sofa or chair.  You collect the remains of the heater for further examination to have a lab determine if the heater switch was in the ``on'' position.  Now you need to start testing your hypothesis.  Is everything consistent with the other fire damage?

There are different damage patterns behind the sofa.  On the right side there is fire damage on the wall behind the right side back cushion of the sofa. The damaged area is distinct from the other fire damage that exists on the wall behind the sofa. There are pieces of charred newspaper near the right arm of the sofa.  The burn mark on the top of the wood coffee table looks like an arrow that is pointing toward the corner.  The char on the legs of the coffee table is very directional and consistent with the heat coming from the corner.  Looking at the inside area of the sofa, you see that there is charred wood on the inside surface of the sofa frame while the outer portions of the wood frame are not charred.  This would indicate that the fire was inside the sofa at some point.  Further, it appears that all of the foam cushioning is burnt off of the right side of the sofa, while there is a piece of polyurethane foam cushion on the left side of the sofa.  However that foam could have gotten moved during the fire suppression.

The chair has more unburned foam cushioning then the sofa does and the seat cushion is still in place with it's distinctive shape, so you know that cushion has not been moved.  The cushion was heated from the top down.  Perhaps the construction of the chair slowed the flame spread through the chair from the bottom up, if the heater was the ignition source.  Or the fire started in the sofa and the radiation from the flames and hot gas layer spread the fire to the chair  The heater still looks good as a source however now you have some fire damage that may not be consistent with the heater as the origin.  Besides interviews what else can you do to eliminate the inconsistencies?   The fire investigators' challenge will be reexamined in Chapter IX.

@@

\begin{table}[h]
  \small
  \centering
  \begin{tabular}{|c|c|c|c|} \hline 
Distance from Wall 		&  Flame Height 			  	& RI=5       		& RI=10      		  \\
    					&  Range          				&  (80~mm grid)  	&  (40~mm grid)  	  \\
    (m)					&  (m) 							& (m) 				& (m) 				  \\ 
\hline    0.61			&  0.73 to 1.14         		& 0.88 to 1.02      & 1.01 to 1.24         \\
\hline    0.31			&		-						& 0.93 to 1.28		& 1.00 to 1.23		    \\
\hline    0.15			&		-						& 0.91 to 1.23		& 1.01 to 1.24			\\
\hline	  0.0			&		-						& 1.03 to 1.34		& 1.20 to 1.52   	  \\
\hline
  \end{tabular}
  \caption[Comparison of estimations of flame height from the natural gas instrumented wall experiments and FDS simulations with RIs of 5 and 10.]{Comparison of estimations of flame height from the natural gas instrumented wall experiments and FDS simulations with RIs of 5, and 10.}
  \label{tab:FDS_FH_IWNG}
\end{table}
